---
title: "Impact of severe weather events in the US in 1950-2011"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```

## Synopsys

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database and aims at exploring the correlation between event types and the magnitude of health and economic impact they bear.

## Loading and Processing the Raw Data

The data for this project comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

There is also some documentation of the database available.

* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r download, cache=TRUE}
library(plyr)
library(dplyr)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "storms.bz2")
storms <- tbl_df(read.csv("storms.bz2"))
```

First, let's look at the data

```{r}
head(storms)
```

Formatting the column names

```{r preprocess}
names(storms)<- tolower(make.names(names(storms), unique = T, allow_ = F))
```


### Population damage by event type

The columns we are interested in are "fatalities" and "injuries". The storm database contain only *direct* injuries, in other words the ones that can be directly attributed to the observed weather event. Let's look at the summary of those columns and see if there's any missing data.

```{r}
summary(storms$fatalities)
```

```{r}
summary(storms$injuries)
```

### Economic damage by event type

The most interesting columns here would be "propdmg" and "cropdmg" with their value magnitutes in "propdmgexp" and "cropdmgexp" respectively.

Let's take a look at the property data first

```{r}
summary(storms$propdmg)
```

```{r}
summary(storms$propdmgexp)
```

Unfortunately we can see that not all the numbers are properly formatted as per NWS storm data documentation. In particular, letters "K", "M" and "B" mean 10^3, 10^6 and 10^9 magnifyers respectively, and we can assume that "H/h" would mean hundreds. We can also assume that numbers directly signify magnitude, and "", "-", "0" mean magnitude of 1. Rest of the symbols remain mysterious and demand further investigation, but seeing that they account for relatively small amount of observations we would ignore them for this research and treat them as missing ones.

```{r clean.property, cache=TRUE, warning=FALSE}
storms <- storms %>%
        mutate(propdmg.m = with(., case_when(
        (propdmgexp == "h" | propdmgexp == "H")~2,
        (propdmgexp == "k" | propdmgexp == "K")~3,
        (propdmgexp == "m" | propdmgexp == "M")~6,
        (propdmgexp == "b" | propdmgexp == "B")~9,
        (propdmgexp %in% c("","0","-"))~1,
        (propdmgexp %in% as.character(1:9))~ as.numeric(as.character(propdmgexp)))))
##Calculate the final loss in USD
storms <- storms %>% mutate(propdmg.v = propdmg*10^propdmg.m)
```

We deal similarly with crops damage data

```{r clean.crops, cache=TRUE}
storms <- storms %>%
        mutate(cropdmg.m = with(., case_when(
        (cropdmgexp == "h" | cropdmgexp == "H")~2,
        (cropdmgexp == "k" | cropdmgexp == "K")~3,
        (cropdmgexp == "m" | cropdmgexp == "M")~6,
        (cropdmgexp == "b" | cropdmgexp == "B")~9,
        (cropdmgexp %in% c("","0","-"))~1,
        (cropdmgexp %in% as.character(1:9))~ as.numeric(as.character(cropdmgexp)))))

storms<- storms %>% mutate(cropdmg.v = cropdmg*10^cropdmg.m)
```  

##Results

### Health impact

For the health impact we will look at the top 10 event types that brought most fatalities and injuries to US throughout the history of observations. 

```{r population.bytype, cache=TRUE}
storms.toph <- storms %>%
        group_by(evtype) %>%
        summarize(total = sum(fatalities,injuries)) %>%
        top_n(10) %>%
        arrange (desc(total))
storms.toph
```
We can see that `r storms.toph$evtype[1]` was the most harmful type of event in the explored preiod.

### Economic impact

For the economic impact we are looking into amount of property and crop damage

```{r prop.bytype, cache=TRUE}
storms.econ <- storms %>%
        group_by(evtype) %>%
        summarize(totalprop = sum(propdmg.v), totalcrop = sum(cropdmg.v), total = sum(propdmg.v+cropdmg.v)) %>%
        top_n(10, wt = total) %>%
        arrange (desc(total))
```

```{r}
storms.econ
```

We can see that for crop the most damaging event is `r storms.econ$evtype[which.max(storms.econ$totalcrop)]`

And for property  - `r storms.econ$evtype[which.max(storms.econ$totalprop)]`

However for total economic impact we are looking at overall costs

```{r economy.plot}

library(ggplot2)
ggplot(data=storms.econ, aes(x=reorder(evtype, -total), y=total/1000000000)) + 
        geom_bar(stat="identity") + 
        labs(title="Weather event types associated with most property and crop damage", 
                y="Loss, bln USD", 
                x = "Event Type") + 
        theme(axis.text.x = element_text(angle = 90, vjust=0.5))
```

### Environment data
```{r}
sessionInfo()
```

The full data for this research is available at [GitHub](https://github.com/lenin-grib/weather-events-impact)

